---
unit_title: "Spatial basics"
theme: wrangle
needs: manip-basics
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Spatial packages

In R, there are two main lineages of tools for dealing with spatial data: sp and sf. 

* sp has been around for a while (the first release was in 2005), and it has a
  rich ecosystem of tools built on top of it. However, it uses a rather 
  complex data structure, which can make it challenging to use. 
  
* sf is newer (first released in October 2016) so it doesn't have such a 
  rich ecosystem. However, it's much easier to use and fits in very naturally 
  with the tidyverse. 

## Loading data

To read spatial data in R, use `read_sf()`. The following code reads an example dataset built into the sf package.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(sf)

# The counties of North Carolina
nc <- read_sf(system.file("shape/nc.shp", package = "sf"))
```

Here, `read_sf()` reads from a __shapefile__. Shapefiles are the most common way to store spatial data. Despite the name, a shapefile is actually a collection of files, not a single file. Each file in a shapefile has the same name, but a different extension. Typically, you'll see four files:

* `.shp` contains the geometry 

* `.shx` contains an index into that geometry.

* `.dbf` contains metadata about each geometry (the other columns in the data
  frame).

* `.prf` contains the coordinate system and projection information. You'll 
  learn more about that shortly.
  
`read_sf()` can read in the majority of spatial file formats, and can likely handle your data even if it isn't in a shapefile.

## Data structure

`read_sf()` reads in spatial data and creates a tibble, so `nc` is a tibble.

```{r}
class(nc)
```

`nc` functions like an ordinary tibble with one exception: the __geometry__ column. 

```{r}
nc %>% 
  select(geometry)
```

The `geometry` column contains __simple features__. Simple features are a standard way of representing geometry types. The most basic geometry types are point, linestring, and polygon.

```{r echo=FALSE}
set.seed(357)

polygon_points <- 
  list(rbind(c(1, 1), c(3, 1), c(3, 3), c(1, 3), c(1, 1)))
linestring_points <- matrix(sample(1:5, 8, replace = TRUE), 4, 2)
  
tribble(
  ~type, ~geometry,
  "Point", st_point(x = c(2, 3)),
  "Linestring", st_linestring(linestring_points),
  "Polygon", st_polygon(polygon_points)
) %>% 
  mutate(type = fct_relevel(type, "Linestring", after = 1)) %>% 
  st_as_sf() %>% 
  ggplot() +
  geom_sf() +
  facet_grid(cols = vars(type)) 
```

There are also _multi_ variants of these basic types: multipoint, multilinestring, and multipolygon. These multi types can contain multiple points, linestrings, or polygons. The final type, geometry collection, can contain multiple different types (e.g., points, linestrings, and multipolygons).

You can build up complex spatial visualizations using just these geometry types. A point, for example, could represent a city, a multilinestring could represent a branching river, and a multipolygon could represent a country composed of different islands. 

`nc`'s `geometry` column contains multipolygons. 

```{r}
nc$geometry
```

You can use `plot()` to visualize the `geometry` column.

```{r nc-plot}
plot(nc$geometry)
```

Each row in `nc` represents a single county. If you pull out a single row and plot, you'll get the shape of a single county. 

```{r}
alleghany_county <- 
  nc %>% 
  filter(CNTY_ID == 1827)

plot(alleghany_county$geometry)
```

`plot()` works, but is limited. In the next unit, you'll learn how to use ggplot2 to create more complex spatial visualizations. 

### Geometry

Let's dive a little deeper into the structure of the `geometry` column. The `geometry` column is a __list column__. You're used to working with tibble columns that are atomic vectors, but columns can also be lists. List columns are incredibly flexible because a list can contain any other type of data structure, including other lists. 

```{r}
typeof(nc$geometry)
```

Let's pull out one piece of the `geometry` column so you can see what's going on under the hood.

```{r}
currituck <- 
  nc %>% 
  filter(NAME == "Currituck")

currituck_geometry <- currituck$geometry[[1]]
```

Currituck county is made up of three separate landmasses.

```{r}
plot(currituck_geometry)
```

How does a single row of the `geometry` column represent these three separate shapes? It turns out that each element of `nc`'s `geometry` column, including `currituck_geometry`, is actually a list of lists of matrices.

```{r}
str(currituck_geometry)
```

The top-level list represents landmasses, and has one element for each landmass in the county. Currituck county has three landmasses, and so has three sub-lists.

In our data, each of these sub-lists only has a single element. 

```{r}
map_int(currituck_geometry, length)
```

These sub-lists would contain multiple elements if they needed to represent a landmass that contained an lake, or a landmass that contained a lake with an island, etc. 
    
Each of these elements is a matrix that gives the locations of points along the polygon boundaries.
    
```{r}
currituck_geometry[[2]][[1]]
```
  
## Manipulating with dplyr

sf objects like `nc` are tibbles, so you can manipulate them with dplyr. The following code finds all counties in North Carolina with an area greater than 2e9 meters squared.

```{r}
nc_large <-
  nc %>%
  mutate(area = st_area(geometry) %>% as.numeric()) %>%
  filter(area > 2e9) 
```

`st_area` finds the area of a geometry, and returns an object with units (in this case, $m^2$).

```{r}
alleghany_county %>% 
  st_area(geometry)
```

These units are helpful, but can be annoying to work with, which is why we used `as.numeric()` in the earlier `mutate()` statement. `as.numeric()` converts the results to a regular numeric vector.

The sf package contains many helpful functions, like `st_area()`, for working with spatial data. They all start with `st_`, so you can type `sf::st_` to see a pop-up list of all the options. The sf package [reference page](https://r-spatial.github.io/sf/reference/index.html) also lists all functions in the package.

## Coordinate system

Earlier, you saw the following matrix.

```{r}
currituck_geometry[[2]][[1]]
```

This matrix gives the position of points along the boundary of one of Currituck county's landmasses. However, these points are pretty meaningless. We need to know the reference point, or __coordinate reference system__ (CRS) of the `nc` data. 

Spatial data is often described in terms of latitude and longitude. The function `st_is_longlat()` returns `TRUE` if the data uses latitude and longitude.  

```{r}
st_is_longlat(nc)
```

Now we know that the points in the matrix are latitudes and longitudes. Unfortunately, this still isn't enough information to figure out exactly which locations on Earth these points refer to. Latitude and longitude values are based on the assumption that the Earth is a smooth ellipsoid, but this assumption isn't true. There are many ways to approximate the shape of the Earth, and different approximations work better for different geographic locations. These different approximations are called the __datum__.

Take two minutes and watch this simple explanation of the datum: https://www.youtube.com/watch?v=xKGlMp__jog

We can use `st_crs()` to find the datum and other CRS metadata of our data.

```{r}
st_crs(nc)
```

The datum of `nc` is "NAD27", the [North American Datum](https://en.wikipedia.org/wiki/North_American_Datum) of 1927 (NAD27).

It's useful to understand the datum and know why it exists, but you don't need to worry too much about it if you're using sf and ggplot2. These packages take care of most of the details for you.

