---
unit_title: "Spatial visualization"
theme: visualize
needs: [spatial-basics]
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Setup

We'll start by reading in a couple of sample data sets, as well as loading the sf package.

```{r setup, message = FALSE}
library(tidyverse)
library(sf)

nc_1974 <- 
  read_sf(system.file("shape/nc.shp", package = "sf")) %>% 
  select(FIPS, NAME, births = BIR74, sids = SID74,  geometry) %>% 
  rename_all(str_to_lower) 

states <- st_as_sf(maps::map("state", plot = FALSE, fill = TRUE))
```

In Spatial basics, you learned how to plot spatial data using the `plot()` function. Now, we'll show you how to visualize spatial data using `geom_sf()`. The advantage of `geom_sf()` is that, as part of ggplot2, you can layer geoms, use scales functions to change colors, tweak your legends, alter the theme, etc.

## `geom_sf()`

`nc_1974` contains data on deaths from Sudden Infant Death Syndrome (SIDS) in 1974 for counties in North Carolina.

```{r}
nc_1974
```

First, let's just plot the county boundaries. `geom_sf()` works like the other geoms, except that you don't need to specify an aesthetic mapping. By default, `geom_sf()` will plot your `geometry` column. 

```{r}
nc_1974 %>% 
  ggplot() +
  geom_sf()
```

Notice that ggplot2 takes care of setting the aspect ratio correctly. If we just plot a single county in North Carolina, the default aspect ratio will be different.

```{r}
nc_1974 %>% 
  filter(name == "Anson") %>% 
  ggplot() +
  geom_sf()
```

Just like with other geoms, you can supply additional aesthetics. For polygons like the counties of North Carolina, `fill` is the most useful aesthetic. Let's visualize the number of births by county.

```{r}
nc_1974 %>% 
  ggplot(aes(fill = births)) +
  geom_sf(color = "white", size = 0.2)
```

Maps like this one, in which geographic areas are colored according to some variable, are called _choropleths_. 

```{r}
nc_1974 %>% 
  ggplot(aes(fill = births)) +
  geom_sf(color = "white", size = 0.2)
```

Notice how it's difficult to tell the difference between many of the counties, especially those with very low numbers of births.

Let's look at the distribution of births. 

```{r}
nc_1974 %>% 
  ggplot(aes(births)) +
  geom_histogram(binwidth = 500) +
  scale_x_continuous(breaks = seq(0, 2e5, 2e3))
```

Almost all of the counties fall between 0 and 8,000, but because there are some counties with births up to 22,000, the sequential color scale will linearly map colors between 0 and 22,000. 

One way to deal with this problem is to cap `births`.

```{r}
nc_1974 %>% 
  mutate(births = pmin(births, 1e4)) %>% 
  ggplot(aes(fill = births)) +
  geom_sf()
```

## Layering

`states` and `nc_1974` each has a different coordinate reference system (CRS).

```{r}
st_crs(states)

st_crs(nc_1974)
```

When you include multiple geospatial layers, ggplot2 will ensure that they all have a common CRS so that it makes sense to overlay them. It will use the CRS of the first layer for all layers.

```{r}
ggplot() +
  geom_sf(data = states) + 
  geom_sf(data = nc_1974)
```

In the above code, we didn't pipe any data into `ggplot()` because each layer uses different data.

Because `geom_sf()` is part of ggplot2, you can combine `geom_sf()` with other geoms.  

```{r}
raleigh <- 
  tibble(
    x = -78.6382,
    y = 35.7796,
    label = "Raleigh"
  )

nc_1974 %>% 
  mutate(births = pmin(births, 1e4)) %>% 
  ggplot() +
  geom_sf(aes(fill = births)) +
  geom_point(aes(x, y), color = "red", data = raleigh) +
  geom_label(aes(x, y, label = label), nudge_x = 0.8, data = raleigh) +
  scale_fill_gradient(low = "#efedf5", high = "#756bb1")
```

In this case, the `x` and `y` positions of the Raleigh dot are assumed be in the same CRS as the sf object.

You can find this lighter color scale, and many others, at the [ColorBrewer website](http://colorbrewer2.org/#).

## `coord_sf()`

You'll need to use `coord_sf()` for two reasons:

* You want to zoom into a specified region of the plot.

  The following plot shows the SIDS rate by county.
  
    ```{r}
    nc_1974 <-
      nc_1974 %>% 
      mutate(sids_rate = sids / births)
    
    nc_1974 %>% 
      ggplot(aes(fill = sids_rate)) + 
      geom_sf()
    ```
  
  Let's zoom in on the county with the highest rate (and also change the color scale).
  
    ```{r}
    nc_1974 %>% 
      ggplot(aes(fill = sids_rate)) + 
      geom_sf() +
      scale_fill_gradient(low = "#efedf5", high = "#756bb1") +
      coord_sf(xlim = c(-80.5, -79), ylim = c(34.8, 36))
    ```
  
* You can also use `coord_sf()` to override the projection in the data. If you don't specify the `crs` argument to `coord_sf()`, if defaults to the CRS given in the data.

    The CRS for the `states` data is

    ```{r}
    states %>% 
      st_crs()
    ```
    
    which produces a map that looks like this:
    
    ```{r}
    states %>% 
      ggplot() +
      geom_sf()
    ```
  
    If you're visualizing the entire US, it's better to use the US Albers projection. Albers is an equal-area projection, which means it accurately reflects the areas of geographic regions. Below is the proj.4 string for the US Albers projection. 
    
    ```{r}
    US_ALBERS <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +datum=WGS84 +no_defs"
    
    states %>% 
      ggplot() +
      geom_sf() +
      coord_sf(crs = US_ALBERS)
    ```